---
title: "session07 - exercises"
format: html
editor: visual
date: 2025-12-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_libraries
#| echo: false

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayesrules)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

The main learning goals of this week's practical parts are:

-   becoming familiar with assessing MCMC diagnostics
-   learning how to address common convergence issues

We will continue analyzing the data from @winter2013making.

```{r}
#| label: preprocessing

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv")

```

We will use a decent version of the latest homework model (see `06_homework_solutions.qmd`). We square-root-transformed the distance measure. The model is saved already, so if you execute this chunk, it should not start sampling.

```{r}
#| label: sqrt-model

sim <- sim |> 
  mutate(sqrt_dist = sqrt(Distance))

# some priors
my_priors <- c(prior(normal(9, 3), class = 'Intercept'),
               prior(normal(0, 3), class = 'b', coef = 'SemSimilar'))

# run model
xmdl_sqrt <- brm(sqrt_dist ~ Sem,
            family = gaussian(),
            file = "../../models/sqrt_sim.RDS",
            prior = my_priors,
            data = sim,
            chains = 4,
            cores = 4, 
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 42)

```

# (A) Assessing chain convergence

Use the `mcmc_plot()` function to assess if the sampling chains have converged. What is your assessment? Do they look like healthy caterpillars.

```{r}
#| label: mcmc-plot

# YOUR CODE HERE
# YOUR ASSESSMENT HERE

```

# (B) Assessing chain convergence

Double check that different chains yields similar posteriors with `mcmc_dens_overlay()` function. What is your assessment?

```{r}
#| label: mcmc-plot

# YOUR CODE HERE
# YOUR ASSESSMENT HERE

```

# (C) Checking convergence diagnostics

Look at the model output and assess whether Rhat, and effective sample sizes (ESS) are satisfying

```{r}
#| label: convergence

# model output
xmdl_sqrt

# Rhat values are ~ 1.00 
# bulk ESS > 400,
# tail ESS > 100

# All good!

```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
