---
title: "session06 - homework - solutions"
format: html
editor: visual
date: 2025-12-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_libraries
#| echo: false

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesrules)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from @winter2013making.

```{r}
#| label: preprocessing

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv")

```

In this study, they investigated the conceptual structure of the metaphor “SIMILARITY IS PROXIMITY.” They did a series of experiments in which they investigated whether participants judge entities to be closer to each other when they are thought to be more similar. They presented participants with descriptions of two cities. These cities either sounded similarly or they sounded differently (`Phon`). Moreover, the descriptions depicted the cities as either semantically similar or different (`Sem`). After reading both descriptions and seeing both names, participant had to draw into a map and had to randomly guess where these two cities are located. The authors measured the `Distance` between the two cities for each participant in millimeters.

# (A) Descriptive statistics and data visualization

Let's visualize the raw `Distance` distribution for both `Sem` categories making a density plot of it and describe the distributions.

```{r}
#| label: show_overall_distribution

sim |> 
  ggplot(aes(x = Distance, fill = Sem)) +
  geom_density() +
  facet_grid(.~Sem)+
  theme_minimal()

# The distributions are somewhat skewed towards the left with a heavy right tale. We will come back to this later

```

# (B) Think about priors

The researchers are interested in whether semantic similarity between the cities affects how close/far apart participants draw the cities, which we can model with the following formula:

```{r}
#| label: model_formula

mdl_formula = bf(Distance ~ Sem)

```

Use the `get_prior()` function and add `mdl_formula` as its first argument and the `sim` dataset as its second. This will give you a list of parameters that the model will estimate and will need a prior for.

```{r}
#| label: get_prior

get_prior(mdl_formula, sim) 

```

Generally, every coefficient of our model corresponds to one parameter under the class `b`, except for the `Intercept` which is its own class. So we have a prior for the `Intercept` of the linear model and a prior for the coefficient `SemSimilar`. Priors of class `b` for our predictors always receive a flat uniform prior by default if you do not specify it otherwise. The default prior for the `Intercept` is data driven and uses a student-t distributions that covers most of the data. We also have a prior for `sigma` which is the residual variance and again is by default assigned a student-t prior informed by the variance in the data.

Now I want you to specify normally distributed weakly informative priors which respect the following knowledge we have: Participants needed to draw the cities on a map on a piece of din-A4 paper. The furthest two points on the map where 214 mm apart.

Your task: First use the `plot_normal()` function to find a normal distribution that covers reasonable Distance values under the above described assumptions about the physical set-up of the experiment.

Ignore for now that you might generate values that are impossible at the tails of the distribution.

```{r}
#| label: prior_exploration

# the distribution should allow all values between 1mm and 214 mms
# if participants use the entire space randomly, we would expect values around 107 to be the most common
# and values that are lower or higher, becoming less common
plot_normal(107, 50)

```

# (C) Specify priors

Now based on your exploration, specify priors. Remember what the two coefficients `Intercept` and `SemSimilar` mean: `Intercept` reflects the average Distance measure for `Sem = Different`, and `SemSimilar` reflects the estimated difference between the Intercept and `Sem = Similar`. Define a prior that assumes normally distributed Distance measures somewhere in the middle of the map for both conditions:

```{r}
#| label: spec_priors

my_priors <- c(prior(normal(107, 50), class = 'Intercept'),
               prior(normal(0, 50), class = 'b', coef = 'SemSimilar'))

```

# (D) Run a prior predictive check

Add your priors to the model below. We will run the model and sample only from our priors, i.e. we ignore the data. This is a good sanity check to see if our priors make sense and what they would predict. Plug in the name of your priors from (C) and run the code chunk.

```{r}
#| label: model_priors

xmdl_prior <- brm(Distance ~ Sem,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            data = sim,
            seed = 42)


# We do a little wrangling to get it into a nice shape
xmdl_prior |> 
  # first extract the samples
  spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem~., ncol = 1)

```

Do the priors reflect what you had in mind? If not, go back to (C) and play around with the priors until you are satisfied.

# (E) Run the full model

If you are happy with your priors, run the full model and plot the posteriors for semantically different and similar cities just like above. How did the posteriors change after accounting for the data?

```{r}
#| label: model_posteriors

xmdl <- brm(Distance ~ Sem,
            family = gaussian(),
            #prior = #YOUR PRIORS HERE
            prior = my_priors,
            data = sim,
            seed = 42)

# We do a little wrangling to get it into a nice shape
xmdl |> 
  # first extract the samples
  spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem~., ncol = 1)

```

After seeing the data, the very diffuse priors have given way to much more narrow and shifted posteriors.

We can also plot the data in an alternative way, since we are mostly interested in the difference between the two conditions:

```{r}
#| label: alternative_plot

# We do a little wrangling to get it into a nice shape
xmdl |> 
  # first extract the samples for the differences only
  spread_draws(b_SemSimilar) |> 
  # plot
  ggplot(aes(x = b_SemSimilar)) +
    geom_density(fill = "purple") 

```

Now, one last step, before we move on. Let's check whether this model could have generated the data using posterior predictive checks with the following function: `pp_check()`.

Throughout our modeling process, we'll constantly use `pp_check()`, which is a way for us to know whether our model could have generated the data. This is a way of investigating whether the model 'fits' well. Posterior predictive checks can tell us, for example, whether the chosen response distribution (in this case Gaussian) was appropriate, or whether the model will have to be expanded, that is, whether particular predictors are missing.

```{r}
#| label: pp_check

pp_check(xmdl, ndraws = 100)

```

The mismatch between the real data (dark blue line) and the simulated data (light blue lines) is not *too* bad, but notice the right tale of the observed data. This might make sense when we consider the constraints of the presented map and the constraints of the task. It might be more intuitive to place the cities somewhat central on the map which results in less Distance between them. Very large differences might be less natural to participants.

So this is a first indication that the model assumptions do not perfectly fit the observed values.

# OPTIONAL (F) fix the assumption violation

Can you fix the mismatch between the observed data (dark blue line) and the simulated data (light blue lines) with, for example, a transformation of the Distance measure?

```{r}
#| label: transform

# the observed data have a gentle right tale, so I chose a gentle transformation like the square root transform
sim <- sim |> 
  mutate(sqrt_dist = sqrt(Distance))

# some priors
my_priors <- c(prior(normal(9, 3), class = 'Intercept'),
               prior(normal(0, 3), class = 'b', coef = 'SemSimilar'))

# run model
xmdl_sqrt <- brm(sqrt_dist ~ Sem,
            family = gaussian(),
            prior = my_priors,
            data = sim,
            seed = 42)

# posterior_predictive check
pp_check(xmdl_sqrt, ndraws = 100)

# looks much better. 

```

# OPTIONAL (G) specify priors that will reduce the effect of SEM to zero

Now I want you to specify priors that will dominate the posterior, i.e. they will eliminate any predicted difference between conditions.

```{r}
#| label: too-strict_priors

# some priors
my_priors2 <- c(prior(normal(9, 3), class = 'Intercept'),
                # we specify the difference between conditions as very close to zero
                prior(normal(0, 0.01), class = 'b', coef = 'SemSimilar'))

# run model
xmdl_sqrt_null <- brm(sqrt_dist ~ Sem,
            family = student(),
            prior = my_priors2,
            data = sim,
            seed = 42)

pp_check(xmdl_sqrt_null, ndraws = 100)

# We do a little wrangling to get it into a nice shape
xmdl_sqrt_null |> 
  # first extract the samples for the differences only
  spread_draws(b_SemSimilar) |> 
  # plot
  ggplot(aes(x = b_SemSimilar)) +
    geom_density(fill = "purple") 

```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
