---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim
```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

#specifying model formula based on hw06 and get prior parameters
mdl_formula = bf(Distance ~ Sem)
get_prior(mdl_formula, sim) 

#set priors
my_priors <- c(prior(normal(107, 50), class = 'Intercept'),
               prior(normal(0, 50), class = 'b', coef = 'SemSimilar'))

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

#YOUR MODEL HERE
#run model sampling only from priors
xmdl_prior <- brm(Distance ~ Sem,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            data = sim,
            seed = 42)
summary(xmdl_prior)

#run model normally
xmdl <- brm(Distance ~ Sem,
            family = gaussian(),
            prior = my_priors,
            data = sim,
            seed = 42)
summary(xmdl)
```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

#YOUR CODE HERE
pp_check(xmdl, ndraws = 100)
# looks good 
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# YOUR CODE HERE
mcmc_plot(xmdl, type = "trace")
# 'no divergences to plot', all looks good
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

#YOUR CODE HERE
# check posterior 
posterior_summary(xmdl)
# research question: 
# b_Intercept: 82.52857, the average Distance in mm that a participant will place two cities apart from each other
# b_SemSimilar: -10.13035, with each increase in unit of Distance, the degree of Semantic Similarity decreases by approximatey 10 units
# uncertainty: the 95% CI is between -18.25 and -1.88, average difference -10.13035, 

# visualise posteriors
simp <- xmdl %>% 
  spread_draws(b_SemSimilar) %>% 
  ggplot(aes(x = b_SemSimilar)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   scale_x_continuous(limits = c(-0.1, 0.1)) +
   labs(x = "Semantic Similarity",
        y = "") +
   theme_minimal()
simp

# posterior probabilities of direction: 99.08%, meaning that there is a 99.08% probability that the parameter is strictly negative 
xmdl %>% 
  spread_draws(b_SemSimilar) %>% 
  select(b_SemSimilar) %>% 
  p_direction()

# Bayes Factors: 1.43, meaning that there is not much evidence for the alternative hypothesis that the effect of SemanticSimilarity is at 0 
# extract posteriors
posteriors <- xmdl %>% 
  spread_draws(b_SemSimilar) %>% 
  select(b_SemSimilar)
#refit and get priors
priors <- xmdl_prior %>% 
  spread_draws(b_SemSimilar) %>% 
  select(b_SemSimilar)
#run bayes factor against null value
install.packages("logspline")
bayesfactor_parameters(posteriors, priors, direction = "two-sided", null = 0)
```

## (g) Write up your entire analysis:

Posterior distributions of the model indicate that with every increase in the unit of Distance, Semantic Similarity is estimated to decrease by about 10 units, with a 95% Credible Interval of -18.25 to -1.88.

99.08% of the posterior samples are negative, indicating that positive values are implausible.

A calculation of the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], or Bayes Factor, indicates the degree by which the mass of the posterior distribution has shifted further away from the point null value. The Bayes Factor here is 1.43, indicating merely anecdotal evidence for the alternative hypothesis that the effect of SemanticSimilarity is point null[@lee-wagenmakers_Bayesian_book_2014].

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
